The \tsc{r}-pronoun and postposition \tit{waar mee} in Dutch

### 2021-11-17 paper.tex
sent revised version

\title{The \tsc{r}-pronoun plus postposition \tit{waar mee} in Dutch}
\author{Fenna Bergsma\\Goethe-Universität Frankfurt}
\date{\today}

@@

Dutch has the preposition \tit{met} `with' that expresses the instrumental. In \ref{ex:metdp}, \tit{met} `with' is combined with a full DP. The inanimate pronoun in Dutch is \tit{t} `it'.\footnote{The longer form that is mostly used in writing is \tit{het} `it'. I will use the spoken variant \tit{t} throughout this paper.} \ref{ex:tverb} illustrates that \tit{t} `it' can be used as the object of a verb.

@@

\tit{Met} `with' and \tit{t} `it' do not appear together, as illustrated in \ref{ex:neemett}. Instead, Dutch uses the \tsc{r}-pronoun \tit{r} `there' and the postposition \tit{mee} `with', as shown in \ref{ex:jar mee}.\footnote{The \tsc{r}-pronoun \tit{r} in \ref{ex:jar mee} can be written as \tit{er}, \tit{der} and \tit{r} and pronounced as respectively /ɛr/, /dər/ or /ər/. There is no clear meaning difference between these forms that is relevant for the discussion in this paper
(see \citet{wesseling2018} for discussion).
In my examples I use \tit{r}, but the other forms fit just as well.}$^,$\footnote{In this paper I do not make any claims about the distinction between prefixes and prepositions, or suffixes and postpositions.}

\ex.\label{ex:r meeconst}
\ag. *Ik schilder met t.\\
 I paint with it\\
 `I am painting with it.'\label{ex:neemett}
\bg. Ik schilder r mee.\\
 I paint there with\\
 `I am painting with it.'\label{ex:jar mee}

\tsc{R}-pronouns are nominal elements that are syncretic with locative pronouns, which in Dutch means they contain the morpheme \tit{r} \citep[cf.][]{riemsdijk1978,koopman1994}. The adpositions they combine with obligatorily follow the \tsc{r}-pronoun (compare \ref{ex:jar mee} and \ref{ex:neemeer}).
Notice also that the preposition \tit{met} `with' differs phonologically from the postposition \tit{mee} `with` (see \ref{ex:metdp} and \ref{ex:jar mee}), although they express the same meaning.

@@

The main question I address in this paper is how to correctly rule out \tit{met t} `with it' in \ref{ex:neemett}, and let the \tsc{r}-pronoun plus postposition \tit{r mee} `with it' in \ref{ex:jar mee} appear. I argue that \tsc{r}-pronouns are not something special, but a consequence of regular spellout mechanisms. Just as \citet{riemsdijk1978} I analyze the \tsc{r}-pronoun plus postposition as a type of allomorph of the preposition plus pronoun. The crucial difference in the current approach is that spellout - not the stipulation of a filter - rules out the ungrammaticality of the preposition plus inanimate pronoun.

This paper focuses on the instrumental \tsc{r}-pronoun plus postposition \tit{r mee} `with it' in Dutch. This instance is interesting for two reasons. First, just as for all \tsc{r}-pronouns, the \tsc{r}-pronoun is syncretic with the locative. In many of the \tsc{r}-pronouns and postpositions, the meaning component of the locative within the \tsc{r}-pronoun plus postposition is intuitive, as many adpositions express locations, directions, etc. However, an instrumental expresses an instrument, which does not have a meaning component associated with location. Ideally, a formal analysis would treat the syncretism as unaccidental but still allow the locative meaning to be absent.
The second reason for focusing on this particular \tsc{r}-pronoun plus postposition is the form of the adposition. The preposition \tit{met} `with' does not only turn into a postposition, but it also changes into \tit{mee} `with' when it is combined with an \tsc{r}-pronoun. This last observation has so far remained unexplained in the literature on \tsc{r}-pronouns (cf. \citealt{koopman1994,riemsdijk1978,noonan2017dutch,abels2003diss}).

The main generalization is that the instrumental \tsc{r}-pronoun plus postposition \tit{r mee} `with it' takes precedence over the instrumental preposition plus inanimate pronoun \tit{met t} `with it'. I show that a single requirement has to be satisfied: the instrumental object needs to form a proper constituent i.e. a constituent to the exclusion of other features. When this condition is not met, the preposition plus inanimate pronoun appear. This can straightforwardly follow in a system in which spellout targets phrasal constituents: Nanosyntax \citep{starke2009}. I work this idea out capturing the following three observations. First, \tsc{r}-pronouns are syncretic with locatives. Second, inanimate pronouns appear with prepositions, and \tsc{r}-pronouns appear with postpositions. And third, the instrumental preposition and postposition differ in form (\tit{met} vs. \tit{mee}).

This paper is structured as follows. In Section \ref{sec:distribution} I show that \tsc{r}-pronouns plus postpositions and prepositions plus inanimate pronouns are in complementary distribution, depending on whether all features form a proper constituent or not. Besides discussing \tit{r mee} `with it' and \tit{met t} `with it', I introduce the \tit{wh}-pronouns \tit{waarmee} `with what' and \tit{met wat} `with what'. For these forms it is intuitive to see that proper constituency is the critical condition.
In Section \ref{sec:takingapart} I decompose \tit{waar mee} `with what' and \tit{met wat} `with what', and I link the morphemes they consist of to specific syntactic structures. Section \ref{sec:derivation} shows in a derivation how \tit{waar mee} `with what' is derived when all features form a proper constituent. I also show how \tit{met wat} `with what' appears when the features do not form a proper constituent. Section \ref{sec:conclusion} concludes. All examples in this paper are from spoken Dutch, unless indicated otherwise. Examples without a reference are constructed and have been verified by native speakers.

@@

The goal of this section is to show that \tsc{r}-pronouns and postpositions appear when all relevant features form a proper constituent, i.e. that all features form a constituent to the exclusion of any other features. I start the section by showing that \tit{r mee} `with it' and \tit{waar mee} `with what` are the default, which is to say that they normally appear and \tit{met t} `with it' and \tit{met wat} `with what' do not. There is one exception, in which \tit{met wat} `with what` has to be used and \tit{waar mee} `with what` cannot, which is when case the relevant features do not form a proper constituent.

@@

\tit{Met} is not the only preposition this occurs with. For example, \tit{op} `on' and \tit{in} `in' cannot combine with the inanimate pronoun \tit{t} either, and the \tsc{r}-pronoun plus postposition are obligatorily used instead.

@@

The situation of the inanimate \tsc{wh}-pronouns resembles the one of the inanimate personal pronouns. \tit{Wat} `what' can function as the object of a verb (see \ref{ex:wat}), but not as the object of a preposition \ref{ex:metwat}. The \tsc{r}-pronoun plus postposition \tit{waar mee} `with what' appears, as shown in \ref{ex:waar mee}.\footnote{The sentence in \ref{ex:metwat} is unacceptable with neutral intonation. It becomes acceptable if \tit{wat} `what' is stressed, for example in a context in which the speaker is highly surprised about the choice of the object the hearer is painting with.}

\ex.
\ag. Wat zie jij?\\
 what see you\\
 `What do you see?'\label{ex:wat}
\bg. *Met wat schilder jij?\\
 with what paint you\\
 `What are you painting with?'\label{ex:metwat}
\bg. Waar mee schilder jij?\\
 where with paint you\\
 `What are you painting with?'\label{ex:waar mee}

\tit{Waar mee} `with what' and not \tit{met wat} `with what' does not only appear in \tsc{wh}-questions, but also in other contexts. \ref{ex:headed} gives an example of a headed relative, and \ref{ex:headless} shows a free relative in which both predicates combine with an instrumental object. The use of \tit{met wat} `with what' is ungrammatical in both contexts, and \tit{waar mee} `with what' is used instead.\footnote{
In this example, \tit{waar} `where' and \tit{mee} `with' are the two left-most elements of the relative clause. It is also possible for \tit{mee} `with' to appear more to the right in the clause and for \tit{waar} to appear as the left-most element on its own.

@@

I return to this variant in Section \ref{sec:derivation}.
}

@@

In sum, the inanimate pronouns \tit{t} `it' and \tit{wat} `what' do not combine with prepositions. They are substituted by, respectively, the \tsc{r}-pronouns \tit{r} `there' and \tit{waar} `where'.

@@

A mismatching free relative is a free relative construction in which the two predicates (the one in the main clause and the one in the relative clause) combine with two different cases. In other words, the case requirements of the two clauses do not match. First consider \ref{ex:gekochtwaarmee}. The predicate in the relative clause, \tit{schildert} `paint', combines with an instrumental object. The predicate in the main clause, \tit{gekocht} `bought' combines with an accusative object. The \tsc{r}-pronoun plus postposition \tit{waar mee} `with what' is used here, shown in \ref{ex:gekochtwaarmee}. The use of \tit{met wat} `with what' is ungrammatical in this context, illustrated in \ref{ex:gekochtmetwat}.

@@

Table \ref{tbl:distribution} summarizes the pattern. When the main clause predicate combines with an accusative and the relative clause predicate with an instrumental, \tit{waar mee} `with what' is grammatical and \tit{met wat} `with what' is ungrammatical. When the main clause predicate combines with an instrumental and the relative clause predicate with an accusative, \tit{waar mee} `with what' is ungrammatical and \tit{met wat} `with what' appears instead.

@@

Below I repeat the examples with instrumental objects discussed in this paper so far. I placed square brackets around the relative clauses in \ref{ex:const3} and \ref{ex:const4}. In each of these examples the instrumental object forms a proper constituent at a certain point in the derivation.

@@

In \ref{ex:mismatchwaarmee}, I repeat the mismatching free relative in which \tit{waar mee} `with what' appears. The predicate \tit{schildert} `paints' combines in the relative clause with the instrumental object. I placed square brackets around the relative clauses in \ref{ex:mismatchwaarmee}. The instrumental object forms a proper constituent within the relative clause, and it can be realized as the \tsc{r}-pronoun plus postposition \tit{waar mee} `with what'.

@@

Consider now the other kind of mismatch. The relative clause predicate \tit{gekocht} `bought' combines with an accusative object. I placed square brackets around the relative clauses in \ref{ex:mismatchmetwat}. The accusative \tsc{wh}-object of a verb is always \tit{wat} `what', as I have shown earlier in \ref{ex:wat}. The instrumental only comes into the picture in the main clause, when \tit{schilder} `paint' combines with an instrumental object. At no point in the derivation does the instrumental object form a proper constituent. \tit{Waar mee} `with what' cannot surface, and \tit{met wat} appears instead.

@@

\ref{ex:summaryconst} summarizes the main results of this section. \tit{Met wat} `with what' can not surface when the features that correspond to \tit{met wat} form a proper constituent. It is always \tit{waar mee} `with what' that appears, as shown in \ref{ex:waar meefr}. \tit{Met wat} `with what' can appear when the instrumental object does not form a proper constituent.
This can be either when \tit{wat} `what' takes a complement before it combines with \tit{met} `with', as shown in \ref{ex:metwatx}, corresponding to the examples in \ref{ex:moremetwat}. The other option is that \tit{wat} `what' is part of a clause that \tit{met} `with' is not a part of, schematized as \ref{ex:metwatfr} and corresponding to \ref{ex:mismatchmetwat}.

@@

In this section I decompose \tit{waar mee} `with what' and \tit{met wat} `with what' into morphemes. I connect these morphemes to parts of syntactic structure. \tit{Waar mee} `with what' and \tit{met wat} `with what' spell out the same set of features, but the distribution is different. I decompose \tit{waar mee} and \tit{met wat} as in \ref{ex:decompose}. Earlier works have decomposed these pronouns in similar ways (cf. \citealt{hachem2015,noonan2017dutch,wesseling2018}).

@@

In this section I first identify \tit{w} and \tit{a(a)} as morphemes that appear in both expressions. Putting these two aside, I then concentrate on \tit{r mee} `with it' and \tit{met t} `with it'.
I propose an account that makes the ungrammaticality of \tit{met t} and the appearance of \tit{r mee} follow from spellout. The analysis accounts for the following three observations, taking \tit{met t} as the point of departure. First, \tit{met} `with' changes from being a preposition to being a postposition. This process is restricted to inanimate pronouns, and it does not apply to full noun phrases and animate pronouns. Second, \tit{met} changes form to \tit{mee}. Third, \tit{t} is replaced by \tit{r}, a morpheme that is associated with the locative in Dutch.
Along the way I introduce some necessary theoretical background on lexicalization in Nanosyntax.

@@

Let me start with the morphemes \tit{w} and \tit{a(a)} that appear in both \tit{waar mee} `with what' and \tit{met wat} `with what'. I assume that they correspond to the same syntactic structure in both. As I am interested in the differences between the two expressions, I do not discuss the featural content of \tit{w} and \tit{a(a)} in depth.

For \tit{w} I follow \posscitet{hachem2015} work on \tit{d} and \tit{w} elements in German and Dutch. In her work, \tit{d} establishes a definite reference and \tit{w} triggers the construction of a set of alternatives in the sense of \citet{rooth1992} (see \citealt{hachem2015} for discussion).\footnote{Throughout the paper, ⇔ indicates the pairing between a lexical tree and a phonological form in a lexical entry, and ⇒ indicates how a node in the syntactic structure is spelled out.}

@@

I follow several authors (cf. \citealt{lander2016,noonan2017dutch,wesseling2018}) in the assumption that the morpheme \tit{a(a)} is related to deixis. Dutch distinguishes between proximal by using \tit{ie} (/i:/) and \tit{i} (/ɪ/) and distal by using \tit{aa} (/aː/) and \tit{a} (/ɑ/), illustrated in \ref{ex:deixis}.\footnote{
\tsc{Wh}-elements combine with the distal marker \tit{a(a)}, and they cannot with the proximal marker \tit{i/ie}. Conceptually, this can be understood if spatial deixis is connected to discourse deixis \citep[cf.][]{colasanti2019}. Proximal is spatially near the speaker and refers to knowledge that the speaker possesses. Distal is spatially away from the speaker and refers to knowledge that the speaker does not possess. In \tsc{wh}-pronouns, the speaker does not have the knowledge, so the distal is used and not the proximal.
}
I assume that the changes from from /ɪ/ into /i:/ and /ɑ/ into /aː/ is due to phonology, possibly as a result of the final /r/.\footnote{I leave the description of the exact phonological process for future research.}\footnote{
Note that in the proximal the initial \tit{d} is also replaced by an \tit{h}. This could indicate that (1) there must be some difference between \tit{hier} `here' and \tit{daar} `there' besides proximal and distal or (2) there is a different distribution between \tit{h} plus \tit{ie} and \tit{d} plus \tit{i} and the difference in vowels is not a matter of phonology. As this is not the focus of this paper, I put this aside for now.
}

@@

 The morpheme \tit{t} can act as nominative, accusative and dative, so it spells out \tit{k}1, \tsc{k}2 and \tsc{k}3.

@@

In addition to the pronominal and case features, I assume that the \tit{t} contains the ontological category \tsc{thing} \citep{kayne2005}.  For reasons of space I will assume singular to be the absence of number and neuter the absence of gender.

@@

 In other words, a lexically stored structure does not have to be identical to the syntactic structure it spells out. The crucial requirement is that the syntactic structure is contained within the lexically stored tree. This has as a consequence that the lexical entry in \ref{ex:entryt} can also be inserted in \ref{ex:tacc} and \ref{ex:tnom}.

 \ex.
 \a. \begin{forest} boom
 [\tsc{accP}
     [\tsc{k2}]
     [\tsc{nomP}
         [\tsc{k1}]
         [ΣP
             [Σ]
             [\tsc{thingP}
                 [\tsc{thing}, roof]
             ]
         ]
     ]
 ]
 {\draw (.east) node[right]{⇔ \tit{t}}; }
 \end{forest}\label{ex:tacc}
 \b. \begin{forest} boom
 [\tsc{nomP}
     [\tsc{k1}]
     [ΣP
         [Σ]
         [\tsc{thingP}
             [\tsc{thing}, roof]
         ]
     ]
 ]
 {\draw (.east) node[right]{⇔ \tit{t}}; }
 \end{forest}\label{ex:tnom}

@@

Notice here that, via the Superset Principle, \tit{r} can also realize the feature \tsc{thing}, as it is contained in \tsc{locationP}. Moreover, for a syntactic structure like \ref{ex:rthing} the lexical entry \ref{ex:entryr} will be inserted and not \ref{ex:entryt}.

@@

Under this analysis, the caseless base form of the neuter singular pronoun in Dutch is actually \tit{r} and \tit{t} is a suppletive nominative, accusative and dative form. The base form only shows up in the higher cases, from instrumental on. I show this in Table \ref{tbl:dutchcases}.

@@

A similar pattern appears in Iron Ossetic, shown in Table \ref{tbl:ossetic}. In the first person singular of this language, it is only the nominative that is suppletive form (\tit{æz}). The higher cases have the stem \tit{mæn} and they combine with the suffixes that nouns normally combine with.

\begin{table}[ht]
	\center
	\caption {Partial paradigm of Iron Ossetic 1\tsc{sg} and noun \citep{erschler2012}}
		\begin{tabular}{lll}
		\toprule
              & 1\tsc{sg}  & head    \\
		\midrule
    \tsc{nom} & æz          & sær-∅   \\
    \tsc{acc} & mæn-∅       & sær-∅   \\
    \tsc{dat} & mæn-æn      & sær-æn  \\
    \tsc{ins} & mæn-æj      & sær-æj  \\
    \bottomrule
\end{tabular}
\label{tbl:ossetic}
\end{table}

@@

\ex.
\ag. bæx-t-imæ æmæ gæl-t-imæ\\
horse-\tsc{pl-com} and ox-\tsc{pl-com}\\\label{ex:horseoxboth}
\bg. bæx-tæ æmæ gæl-t-imæ\\
horse-\tsc{pl} and ox-\tsc{pl-com}\\
`with horses and oxen' \hfill (Iron Ossetic, \citealt[165]{erschler2012})\label{ex:horseoxone}

@@

\ex.\label{ex:suspendedme}
\ag. mæn æmæ Zauyr-æn\\
 1\tsc{sg} and Zaur-\tsc{dat}\\\label{ex:izaurboth}
\bg. *æz æmæ Zauyr-æn\\
 1\tsc{sg} and Zaur-\tsc{dat}\\
 `me and Zaur' \hfill (Iron Ossetic, \citealt[39]{беляев2014} after \citealt{caha2019competition})\label{ex:izaurone}

>>>>

\title{The \tsc{r}-pronoun and postposition \tit{waar mee} in Dutch}
\author{Fenna Bergsma\\Goethe-Universität Frankfurt}
\date{\today}

@@

Dutch has the preposition \tit{met} `with' that expresses the instrumental. In \ref{ex:metdp}, \tit{met} `with' is combined with a full noun phrase. The inanimate pronoun in Dutch is \tit{t} `it'.\footnote{The longer form that is mostly used in writing is \tit{het} `it'. I will use the spoken variant \tit{t} throughout this paper.} \ref{ex:tverb} illustrates that \tit{t} `it' can be used as the object of a verb.

@@

\tit{Met} `with' and \tit{t} `it' do not appear together, as illustrated in \ref{ex:neemett}. Instead, Dutch uses the \tsc{r}-pronoun \tit{r} `there' and the postposition \tit{mee} `with', as shown in \ref{ex:jarmee}.\footnote{The \tsc{r}-pronoun \tit{r} in \ref{ex:jarmee} can be written as \tit{er}, \tit{der} and \tit{r} and pronounced as respectively /ɛr/, /dər/ or /ər/. There is no clear meaning difference between these forms that is relevant for the discussion in this paper
(see \citealt{wesseling2018} for discussion).
In my examples I use \tit{r}, but the other forms fit just as well.}$^,$\footnote{In this paper I do not make a distinction between prefixes and prepositions, or between suffixes and postpositions.}

\ex.\label{ex:r meeconst}
\ag. *Ik schilder met t.\\
 I paint with it\\
 `I am painting with it.'\label{ex:neemett}
\bg. Ik schilder r mee.\\
 I paint there with\\
 `I am painting with it.'\label{ex:jarmee}

\tsc{R}-pronouns are nominal elements that are syncretic with locative pronouns, which in Dutch means they contain the morpheme \tit{r} \citep[cf.][]{riemsdijk1978,koopman1994}. The adpositions they combine with obligatorily follow the \tsc{r}-pronoun (compare \ref{ex:jarmee} and \ref{ex:neemeer}).
Notice also that the preposition \tit{met} `with' differs phonologically from the postposition \tit{mee} `with` (see \ref{ex:metdp} and \ref{ex:jarmee}), although they express the same meaning.

@@

The main question I address in this paper is how to correctly rule out \tit{met t} `with it' in \ref{ex:neemett}, and let the \tsc{r}-pronoun plus postposition \tit{r mee} `with it' in \ref{ex:jarmee} appear. I argue that \tsc{r}-pronouns are not something special, but a consequence of regular spellout mechanisms. Just as \citet{riemsdijk1978}, I analyze the \tsc{r}-pronoun plus postposition as a type of allomorph of the preposition plus pronoun. The crucial difference in the current approach is that spellout - not the stipulation of a filter - rules out the ungrammaticality of the preposition plus inanimate pronoun.

This paper focuses on the instrumental \tsc{r}-pronoun plus postposition \tit{r mee} `with it' in Dutch. This instance is interesting for two reasons. First, just as for all \tsc{r}-pronouns, the \tsc{r}-pronoun is syncretic with the locative. In many of the \tsc{r}-pronouns and postpositions, the meaning component of the locative within the \tsc{r}-pronoun plus postposition is intuitive, as many adpositions express locations, directions, etc. However, an instrumental expresses an instrument, which does not have a meaning component associated with location. Ideally, a formal analysis would treat the syncretism as unaccidental but still allow the locative meaning to be absent.
The second reason for focusing on this particular \tsc{r}-pronoun plus postposition is the form of the adposition. The preposition \tit{met} does not only turn into a postposition, but it also changes into \tit{mee} when it is combined with an \tsc{r}-pronoun. This last observation has so far remained unexplained in the literature on \tsc{r}-pronouns (cf. \citealt{riemsdijk1978,koopman1994,abels2003diss,noonan2017dutch}).

The main generalization is that the instrumental \tsc{r}-pronoun plus postposition \tit{r mee} `with it' by default takes precedence over the instrumental preposition plus inanimate pronoun \tit{met t} `with it'. I show that a single requirement has to be satisfied: the instrumental object needs to form a proper constituent i.e. a constituent to the exclusion of other features. When this condition is not met, the preposition plus inanimate pronoun appear after all. This follows straightforwardly in a system in which spellout targets phrasal constituents: Nanosyntax \citep{starke2009}. I work out this idea capturing the following three observations. First, \tsc{r}-pronouns are syncretic with locatives. Second, inanimate pronouns appear with prepositions, and \tsc{r}-pronouns appear with postpositions. And third, the instrumental preposition and postposition differ in form (\tit{met} vs. \tit{mee}).

This paper is structured as follows. In Section \ref{sec:distribution} I show that \tsc{r}-pronouns plus postpositions and prepositions plus inanimate pronouns are in complementary distribution, depending on whether all features form a proper constituent or not. Besides discussing \tit{r mee} `with it' and \tit{met t} `with it', I introduce the \tsc{wh}-pronouns \tit{waarmee} `with what' and \tit{met wat} `with what'. Examples with these forms allow me illustrate that proper constituency is the critical condition.
In Section \ref{sec:takingapart} I decompose \tit{waar mee} and \tit{met wat}, and I link the morphemes they consist of to specific syntactic structures. Section \ref{sec:derivation} shows the derivation of \tit{waar mee} when all relevant features form a proper constituent and the derivation of \tit{met wat} when the features do not. Section \ref{sec:conclusion} concludes.

All examples in this paper are from spoken Dutch, unless indicated otherwise. Examples without a reference are constructed and have been verified by native speakers.

@@

The goal of this section is to show that \tsc{r}-pronouns and postpositions appear when all relevant features form a proper constituent, i.e. that all features form a constituent to the exclusion of any other features. I start the section by showing that \tit{r mee} `with it' and \tit{waar mee} `with what` are the default, which is to say that they normally appear (and \tit{met t} `with it' and \tit{met wat} `with what' do not). There is one exception, in which \tit{met wat} `with what` has to be used (and \tit{waar mee} `with what` cannot), which is when the relevant features do not form a proper constituent.

@@

\tit{Met} is not the only preposition this occurs with. For example, \tit{op} `on' and \tit{in} `in' cannot combine with the inanimate pronoun \tit{t} `it' either, and the \tsc{r}-pronoun plus postposition are obligatorily used instead.

@@

The situation of the inanimate \tsc{wh}-pronouns resembles the one of the inanimate personal pronouns. \tit{Wat} `what' can function as the object of a verb (see \ref{ex:wat}), but not as the object of a preposition \ref{ex:metwat}. The \tsc{r}-pronoun plus postposition \tit{waar mee} `with what' appears instead, as shown in \ref{ex:waarmee}.\footnote{The sentence in \ref{ex:metwat} is unacceptable with neutral intonation. It becomes acceptable if \tit{wat} `what' is stressed, for example in a context in which the speaker is highly surprised about the choice of the object the hearer is painting with.}

\ex.
\ag. Wat zie jij?\\
 what see you\\
 `What do you see?'\label{ex:wat}
\bg. *Met wat schilder jij?\\
 with what paint you\\
 `What are you painting with?'\label{ex:metwat}
\bg. Waar mee schilder jij?\\
 where with paint you\\
 `What are you painting with?'\label{ex:waarmee}

\tit{Waar mee} `with what' and not \tit{met wat} `with what' does not only appear in \tsc{wh}-questions, but also in other contexts. \ref{ex:headed} gives an example of a headed relative, and \ref{ex:headless} shows a free relative in which both predicates combine with an instrumental object. The use of \tit{met wat} `with what' is ungrammatical in both contexts, and \tit{waar mee} `with what' is used instead.\footnote{
In this example, \tit{waar} and \tit{mee} are the two left-most elements of the relative clause. It is also possible for \tit{mee} to appear more to the right in the clause and for \tit{waar} to appear as the left-most element on its own.

@@

I return to this variant in Section \ref{sec:waarmee}.
}

@@

In sum, the inanimate pronouns \tit{t} `it' and \tit{wat} `what' do not combine with prepositions. They are substituted by the \tsc{r}-pronouns \tit{r} `there' and \tit{waar} `where' and combine with postpositions.

@@

A mismatching free relative is a free relative construction in which the two predicates (the one in the main clause and the one in the relative clause) require arguments with different cases. In other words, the case requirements of the two clauses do not match. First consider \ref{ex:gekochtwaarmee}. The predicate in the relative clause, \tit{schildert} `paint', combines with an instrumental object. The predicate in the main clause, \tit{gekocht} `bought' combines with an accusative object. The \tsc{r}-pronoun plus postposition \tit{waar mee} `with what' is used here, shown in \ref{ex:gekochtwaarmee}. The use of \tit{met wat} `with what' is ungrammatical in this context, illustrated in \ref{ex:gekochtmetwat}.

@@

Table \ref{tbl:distribution} summarizes the pattern. When the main clause predicate combines with an accusative and the relative clause predicate with an instrumental, \tit{waar mee} is grammatical and \tit{met wat} is ungrammatical. When the main clause predicate combines with an instrumental and the relative clause predicate with an accusative, \tit{waar mee} is ungrammatical and \tit{met wat} appears instead.

@@

Below I repeat the examples with instrumental objects discussed in this paper so far, except for the mismatching free relatives. I place square brackets around the relative clauses in \ref{ex:const3} and \ref{ex:const4}. In each of these examples the instrumental object forms a proper constituent at a certain point in the derivation.

@@

In \ref{ex:mismatchwaarmee}, I repeat the mismatching free relative in which \tit{waar mee} `with what' appears. The predicate \tit{schildert} `paints' combines in the relative clause with the instrumental object. I place square brackets around the relative clauses in \ref{ex:mismatchwaarmee}. The instrumental object forms a proper constituent within the relative clause, and it can be realized as the \tsc{r}-pronoun plus postposition \tit{waar mee} `with what'.

@@

Consider now the other kind of mismatch. The relative clause predicate \tit{gekocht} `bought' combines with an accusative object. I place square brackets around the relative clauses in \ref{ex:mismatchmetwat}. The accusative \tsc{wh}-object of a verb is always \tit{wat} `what', as I have shown earlier in \ref{ex:wat}. The instrumental only comes into the picture in the main clause, when \tit{schilder} `paint' combines with an instrumental object. At no point in the derivation does the instrumental object form a proper constituent. \tit{Waar mee} `with what' cannot surface, and \tit{met wat} appears instead.

@@

\ref{ex:summaryconst} summarizes the main results of this section. \tit{Met wat} can not surface when the features that correspond to \tit{met wat} form a proper constituent. It is always \tit{waar mee} that appears, as shown in \ref{ex:waar meefr}. \tit{Met wat} can appear when the instrumental object does not form a proper constituent.
This can be when \tit{wat} takes a complement before it combines with \tit{met}, as shown in \ref{ex:metwatx}, corresponding to the examples in \ref{ex:moremetwat}. The other option is that \tit{wat} is part of a clause that \tit{met} is not a part of, schematized as \ref{ex:metwatfr} and corresponding to \ref{ex:mismatchmetwat}.

@@

In this section I decompose \tit{waar mee} `with what' and \tit{met wat} `with what' into morphemes. I connect these morphemes to parts of syntactic structure. \tit{Waar mee} and \tit{met wat} spell out the same set of features, but the distribution is different. I decompose \tit{waar mee} and \tit{met wat} as in \ref{ex:decompose}. Earlier works have decomposed these pronouns in similar ways (cf. \citealt{hachem2015,noonan2017dutch,wesseling2018}).

@@

In this section I first identify \tit{w} and \tit{a(a)} as morphemes that appear in both expressions. Putting these two aside, I then concentrate on \tit{r mee} and \tit{met t}.
I propose an account that makes the ungrammaticality of \tit{met t} and the appearance of \tit{r mee} follow from spellout. The analysis accounts for the following three observations, taking \tit{met t} as the point of departure. First, \tit{met} changes from being a preposition to being a postposition. This process is restricted to inanimate pronouns, and it does not apply to full noun phrases and animate pronouns. Second, \tit{met} changes form to \tit{mee}. Third, \tit{t} is replaced by \tit{r}, a morpheme that is associated with the locative in Dutch.
Along the way I introduce some necessary theoretical background on lexicalization in Nanosyntax.

@@

Let me start with the morphemes \tit{w} and \tit{a(a)} that appear in both \tit{waar mee} `with what' and \tit{met wat} `with what'. I assume that they correspond to the same syntactic structure in both expressions, As I am interested in the differences between the two expressions, I do not discuss the featural content of \tit{w} and \tit{a(a)} in depth.

For \tit{w}, I follow \posscitet{hachem2015} work on \tit{d} and \tit{w} elements in German and Dutch. In her work, \tit{d} establishes a definite reference and \tit{w} triggers the construction of a set of alternatives in the sense of \citet{rooth1992} (see \citealt{hachem2015} for discussion).\footnote{Throughout the paper, ⇔ indicates the pairing between a lexical tree and a phonological form in a lexical entry, and ⇒ indicates how a node in the syntactic structure is spelled out.}

@@

I follow several authors (cf. \citealt{lander2016,noonan2017dutch,wesseling2018}) in the assumption that the morpheme \tit{a(a)} is related to deixis. Dutch distinguishes between proximal by using \tit{ie} (/i:/) and \tit{i} (/ɪ/) and distal by using \tit{aa} (/aː/) and \tit{a} (/ɑ/), illustrated in \ref{ex:deixis}.\footnote{
\tsc{Wh}-elements combine with the distal marker \tit{a(a)}, and they cannot with the proximal marker \tit{i/ie}. Conceptually, this can be understood if spatial deixis is connected to discourse deixis \citep[cf.][]{colasanti2019}. Proximal is spatially near the speaker and refers to knowledge that the speaker possesses. Distal is spatially away from the speaker and refers to knowledge that the speaker does not possess. In \tsc{wh}-pronouns, the speaker does not have the knowledge, so the distal is used and not the proximal.
}
I assume that the changes from from /ɪ/ into /i:/ and /ɑ/ into /aː/ is due to phonology, possibly as a result of the final /r/.\footnote{I leave the description of the exact phonological process for future research.}\footnote{
Note that in the proximal the initial \tit{d} is also replaced by an \tit{h}. This could indicate that (1) there must be some difference between \tit{hier} `here' and \tit{daar} `there' besides proximal and distal, or (2) there is a different distribution between \tit{h} plus \tit{ie} and \tit{d} plus \tit{i} and the difference in vowels is not a matter of phonology. As this is not the focus of this paper, I put this aside for now.
}

@@

 The morpheme \tit{t} can act as nominative, accusative and dative, so it spells out \tsc{k}1, \tsc{k}2 and \tsc{k}3.

@@

In addition to the pronominal and case features, I assume that the \tit{t} contains the ontological category \tsc{thing} \citep{kayne2005}.  For reasons of space I will assume singular to be the absence of number and inanimate or neuter the absence of gender.

@@

 In other words, a lexically stored tree does not have to be identical to the syntactic structure it spells out. The crucial requirement is that the syntactic structure is contained within the lexically stored tree. This has as a consequence that the lexical entry in \ref{ex:entryt} can also be inserted in \ref{ex:tacc} and \ref{ex:tnom}.

 \ex.
 \a. \begin{forest} boom
 [\tsc{accP}
     [\tsc{k2}]
     [\tsc{nomP}
         [\tsc{k1}]
         [ΣP
             [Σ]
             [\tsc{thingP}
                 [\tsc{thing}, roof]
             ]
         ]
     ]
 ]
 {\draw (.east) node[right]{⇒ \tit{t}}; }
 \end{forest}\label{ex:tacc}
 \b. \begin{forest} boom
 [\tsc{nomP}
     [\tsc{k1}]
     [ΣP
         [Σ]
         [\tsc{thingP}
             [\tsc{thing}, roof]
         ]
     ]
 ]
 {\draw (.east) node[right]{⇒ \tit{t}}; }
 \end{forest}\label{ex:tnom}

@@

Notice here that, via the Superset Principle, \tit{r} can also realize the feature \tsc{thing}, as it is contained in \tsc{locationP}. Moreover, for a syntactic structure as \ref{ex:rthing} the lexical entry \ref{ex:entryr} will be inserted and not \ref{ex:entryt}.

@@

Under this analysis, the caseless base form of the inanimate singular pronoun in Dutch is actually \tit{r}, and \tit{t} is a suppletive nominative, accusative and dative form. The base form only shows up in the higher cases, from the instrumental on. I show this in Table \ref{tbl:dutchcases}.

@@

A similar pattern appears in Ossetic, shown in Table \ref{tbl:ossetic}. In the first person singular of this language, it is only the nominative that is suppletive: \tit{æz}. The higher cases have the stem \tit{mæn} and they combine with the suffixes that nouns normally combine with.

\begin{table}[ht]
	\center
	\caption {Partial paradigm of Ossetic 1\tsc{sg} and noun \citep{erschler2012}}
		\begin{tabular}{lll}
		\toprule
              & 1\tsc{sg}  & head    \\
		\midrule
    \tsc{nom} & æz          & sær-∅   \\
    \tsc{acc} & mæn-∅       & sær-∅   \\
    \tsc{dat} & mæn-æn      & sær-æn  \\
    \tsc{ins} & mæn-æj      & sær-æj  \\
    \bottomrule
\end{tabular}
\label{tbl:ossetic}
\end{table}

@@

\ex.
\ag. bæx-t-imæ æmæ gæl-t-imæ\\
horse-\tsc{pl-com} and ox-\tsc{pl-com}\\\label{ex:horseoxboth}
\bg. bæx-tæ æmæ gæl-t-imæ\\
horse-\tsc{pl} and ox-\tsc{pl-com}\\
`with horses and oxen' \hfill (Ossetic, \citealt[165]{erschler2012} after \citealt{caha2019competition})\label{ex:horseoxone}

@@

\ex.\label{ex:suspendedme}
\ag. mæn æmæ Zauyr-æn\\
 1\tsc{sg} and Zaur-\tsc{dat}\\\label{ex:izaurboth}
\bg. *æz æmæ Zauyr-æn\\
 1\tsc{sg} and Zaur-\tsc{dat}\\
 `me and Zaur' \hfill (Ossetic, \citealt[39]{беляев2014} after \citealt{caha2019competition})\label{ex:izaurone}
